---
title: "Clustering analysis of Retinal Bipolar Cell Drop-seq data"
output:
  html_document: default
  pdf_document: default
---


```{r, warning=FALSE, message=FALSE}
library(Seurat)
library(dplyr)
library(Matrix)
library(sva)
```

##Data preprocessing

### Load the data file containing the expression matrix `bipolar_dge`

```{r, cache.lazy=FALSE, tidy=TRUE,  tidy.opts=list(width.cutoff=80)}
# read in the sparse matrix
bipolar_dge = readMM("data/bp_3k_cells.sparse.matrix")
# read in the gene names (row names)
gene_names = readLines("data/bp_3k_cells.genes.txt")
barcode_names = readLines("data/bp_3k_cells.barcodes.txt")
# assign row and column names for our sparse matrix
rownames(bipolar_dge) = gene_names
colnames(bipolar_dge) = barcode_names
```


### Set up Seurat object
```{r}
bp = CreateSeuratObject(raw.data = bipolar_dge, min.cells = 3)
```

### QC and pre-processing

```{r qc}

mito.genes <- grep(pattern = "^mt-", x = rownames(x = bp@data), value = TRUE)
percent.mito <- Matrix::colSums(bp@raw.data[mito.genes, ]) / Matrix::colSums(bp@data)

# AddMetaData adds columns to object@data.info, and is a great place to stash QC stats
bp <- AddMetaData(object = bp, metadata = percent.mito, col.name = "percent.mito")

# We filter out cells that have unique gene counts over 2,500 or less than 200
# Note that low.thresholds and high.thresholds are used to define a 'gate'
# -Inf and Inf should be used if you don't want a lower or upper threshold.
bp <- FilterCells(object = bp, subset.names = c("percent.mito"), low.thresholds = c(-Inf), high.thresholds = c(0.05))

bp <- FilterCells(object = bp, subset.names = c("nGene"), low.thresholds = c(500), high.thresholds = c(2000))
print(dim(bp@data))
```

### Normalizing the data

```{r normalize}
bp <- NormalizeData(object = bp, normalization.method = "LogNormalize", scale.factor = 1e4)
```

### Detection of variable genes across the single cells


```{r var_genes, message=FALSE, warning=FALSE}
bp <- FindVariableGenes(object = bp, mean.function = ExpMean, dispersion.function = LogVMR, x.low.cutoff = 0.2, x.high.cutoff = 4, y.cutoff = 1.0)

length(x = bp@var.genes)
```
### PCA in prep for tSNE
```{r pca_pre_regress, message=FALSE, warning=FALSE}
bp <- ScaleData(object = bp)
bp <- RunPCA(object = bp, pc.genes = bp@var.genes, do.print = FALSE, pcs.print = 1:2, pcs.compute = 40, maxit = 500, weight.by.var = FALSE)
PCAPlot(object = bp, dim.1 = 1, dim.2 = 2)
```

### tSNE
```{r tsne}
bp <- RunTSNE(object = bp, dims.use = 1:10, do.fast = TRUE)
TSNEPlot(object = bp)
FeaturePlot(bp, features.plot=c('nUMI'))
```




## Batch correction

Bipolar 5 and 6 belong to the second batch, all others are the first batch.

```{r define batches}
batchname = bp@meta.data$orig.ident
batchid = rep(1,length(batchname))
batchid[batchname=="Bipolar5"] = 2
batchid[batchname=="Bipolar6"] = 2
names(batchid) = rownames(bp@meta.data)
bp <- AddMetaData(object = bp, metadata = batchid, col.name = "batchid")
table(bp@meta.data$batchid)
FeaturePlot(bp, features.plot=c('batchid'))
```

```{r saveobject}
save(bp, file = "bp_pre_batch_correct.Robj")
```


### Try regressing out batch via Seurat

```{r, message=FALSE, warning=FALSE}
load("bp_pre_batch_correct.Robj") # restore it 
bp@data = bp@data[Matrix::rowSums(bp@data)>0,]  #after filtering cells, some genes have zero counts
bp <- ScaleData(object = bp, vars.to.regress = c("batchid"))
bp <- RunPCA(object = bp, pc.genes = bp@var.genes, do.print = FALSE, pcs.compute = 40, weight.by.var = FALSE)
bp <- RunTSNE(object = bp, dims.use = 1:10, do.fast = TRUE)
FeaturePlot(bp, features.plot=c('batchid'))
```


### Alternatively, batch-correct using ComBat

```{r, message=FALSE, warning=FALSE}
load("bp_pre_batch_correct.Robj") # restore it before running combat on it
library('sva')
m = as.data.frame(as.matrix(bp@data))
m = m[rowSums(m)>0,]
com = ComBat(m, batchid, prior.plots=FALSE, par.prior=TRUE)
bp@data = Matrix(as.matrix(com))
bp = ScaleData(bp)
bp <- RunPCA(object = bp, pc.genes = bp@var.genes, do.print = FALSE, pcs.print = 1:2, pcs.compute = 40, weight.by.var = FALSE)
bp <- RunTSNE(object = bp, dims.use = 1:10, do.fast = TRUE)
FeaturePlot(bp, features.plot=c('batchid'))
```

